---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---
1. In lecture, we used maximum likelihood and a likelihood ratio test to complete a t-test. We can actually use a likelihood ratio test to compare two models as long as one model is a subset of the other model.
For example, we can ask whether y is a hump-shaped vs. linear function of x by comparing a quadratic
(a + bx + cx^2) vs. linear (a + bx) model. Generate a script that evaluates which model is more appropriate
for the data in data.txt.


```{r}
data2 = read.table("Ex10data.csv", sep =",", header = TRUE )
head(data2)

nllike<-function(p,x,y){ #liklihood function for linear equation
  B0=p[1]
  B1=p[2]
  sigma=exp(p[3])
  expected=B0+B1*x
  nll=-sum(dnorm(x=y,mean=expected,sd=sigma,log=TRUE))
  return(nll)
} 

initialGuess=c(1,1,1) #a spot to start
fit=optim(par=initialGuess,fn=nllike,x=data2$x,y=data2$y)

print(fit)

#Negative log liklihood = 107.7

#Use this function to generate an estimate of the paramaters of best fit for the data and then compare these estimates to coefficient estimates of  a quadradic equation
```


now estimate the parameters for a quadradic model (a + bx + cx^2)
```{r}
nllike2<-function(p,x,y){ #liklihood function for quadradic equation
  #parameters:
  B0=p[1]
  B1=p[2]
  B2=p[3]
  sigma=exp(p[4])
  #deterministic model:
  expected=B0+B1*x+B2*x^2
  #calculate negative log liklihood:
  nll=-sum(dnorm(x=y,mean=expected,sd=sigma,log=TRUE))
  return(nll)
}

initialGuess2=c(1,1,1,1) #a spot to start
fit2=optim(par=initialGuess2,fn=nllike,x=data2$x, x2=(data$x)^2, y=data2$y)

print(fit2)

#The negative log liklihood for this is 128.14 

```

Now calcuate a test statistic using the 2 negative log liklihoods for the 2 models 
then ask if this test stastic falls within the chisquared ditribution
```{r}

# run likelihood ratio test
teststat=2*(fit$value-fit2$value)
teststat
#Does the test statitistic of -40.7 fall within the chisquared ditribution? 

df=length(fit2$par)-length(fit$par)
df

1-pchisq(teststat,df)


```

We got a chisquared value of 1, therfore the more complex model is not significant 


2. A classic model of competition between two species was developed by Lotka & Volterra. This model has two state variables described by two differential equations
??12 < ??11 and ??21 < ??22
Generate a script that uses three or more model simulations to demonstratethe validity of these criteria for
coexistence.

```{r}
### Load the deSolve package and ggplot2 for plotting
library(deSolve)
library(ggplot2)

popmodel = function(t,y,p) {
  #statevariables
  N1=y[1]
  N2=y[2]
  #parameters
  r1=p[1]
  a11=p[2]
  a12=p[3]
  r2=p[4]
  a22=p[5]
  a21=p[6]
  #Model 
  dndt1 = r1*(1 ??? N1*a11 ??? N2*a12)*N1 #a11 is equal to k1, the carrying capacity
  
  dndt2 = r2*(1 ??? N2*a21 ??? N1*a22)*N2 #a22 is equal to k2
  
  return(list(c(dndt1,dndt2)))
  
}
  
# sim 1
times=1:100
N0=c(0.1,0.1)
params2=c(0.5,10,2,0.5,10,0.5) #k is less than alpha for species 1 and greater than alpha for species 2, therfore no coexistence and species 1 outcompetes species 2
sim2=ode(y=N0,times=times,func=popmodel,parms=params2)
out2=data.frame(time=sim2[,1],species1=sim2[,2],species2=sim2[,3])
ggplot(out2,aes(x=time,y=species1))+geom_line()+geom_line(data=out2,mapping=aes(x=time,y=species2),col='red')+theme_classic()


  
```

```{r}
# sim 2
times=1:100
N0=c(0.05,0.3)
params3=c(0.5,10,0.5,0.5,10,0.5)#k is greater than alpha for species 1 AND greater than alpha for species 2, therfore coexistence with species 1 and species 2 occurs
sim3=ode(y=N0,times=times,func=popmodel,parms=params3) 
out3=data.frame(time=sim3[,1],species1=sim3[,2],species2=sim3[,3])
ggplot(out3,aes(x=time,y=species1))+geom_line()+geom_line(data=out3,mapping=aes(x=time,y=species2),col='red')+theme_classic()


```

```{r}
# sim 3
times=1:100
N0=c(0.1,0.1)
params4=c(0.5,10,0.5,0.5,10,2)#k is greater than alpha for species 1 and less than alpha for species 2, therfore coexistence does not occur an species one outcompetes species 2 
sim4=ode(y=N0,times=times,func=popmodel,parms=params4)
out4=data.frame(time=sim4[,1],species1=sim4[,2],species2=sim4[,3])
ggplot(out4,aes(x=time,y=species1))+geom_line()+geom_line(data=out4,mapping=aes(x=time,y=species2),col='red')+theme_classic()
```


Conclusion: 

Sim1: k is less than alpha for species 1 and greater than alpha for species 2, therfore no coexistence and species 1 outcompetes species 2 

Sim 2: k is greater than alpha for species 1 AND greater than alpha for species 2, therfore coexistence with species 1 and species 2 occurs

Sim3: k is greater than alpha for species 1 and less than alpha for species 2, therfore coexistence does not occur an species one outcompetes species 2 